---
title: "STAT8150 Ass2"
format: pdf
editor: visual
---

## Library Import

```{r}
#| include: false
library(car)
library(rjags)
library(ggplot2)
library(dplyr)
#install.packages('ggthemes')
library(ggthemes)
library(tidyr)
library(patchwork)
library(gridExtra)
library(knitr)
library(coda)
```

## Data Import

```{r}
liver_data <- read.csv("cirrhosis.csv")
dim(liver_data)
```

## Data Cleaning

```{r}
#| include: false
# Check missing data
missing_counts <- colSums(is.na(liver_data))
print(missing_counts)
```

```{r}
#| echo: false
# Remove six rows with missing data of stage 
liver_data <- liver_data[!is.na(liver_data$Stage), ]
# Map death status 
liver_data$Death <- ifelse(liver_data$Status == "D", 1, 0)
```

```{r}
#| echo: false
# Check the filtering outcome
dim(liver_data)
```

```{r}
#| echo: false
# Convert the type of stage
liver_data$Stage <- as.numeric(liver_data$Stage)
```

```{r}
#| echo: false
# Convert age (days -> years)
liver_data$Age_years <- liver_data$Age / 365.25
```

```{r}
#| echo: false
# Convert some variables into dummy variables 
# Gender
liver_data$Sex_num <- ifelse(liver_data$Sex == "F", 1, 0)
# Abnormalities 
liver_data$Ascites_num <- ifelse(liver_data$Ascites == "Y", 1, 0)
liver_data$Hepato_num  <- ifelse(liver_data$Hepatomegaly == "Y", 1, 0)
liver_data$Spiders_num <- ifelse(liver_data$Spiders == "Y", 1, 0)
# Edema (in order)
liver_data$Edema_num <- ifelse(liver_data$Edema == "N", 0, 
                               ifelse(liver_data$Edema == "S", 1, 2))
```

```{r}
#| echo: false
# Convert Stage type into ordered factor (1 < 2 < 3 < 4)
liver_data$Stage <- factor(liver_data$Stage, 
                           levels = c(1, 2, 3, 4), 
                           ordered = TRUE)

class(liver_data$Stage)
is.ordered(liver_data$Stage)
```

## Research Question 1

Are the four selected clinical biomarkers statistically significant in predicting sequential cirrhosis stages?

```{r}
# Calculate Spearman's rank correlation for ordinal/continuous variables
# against a target variable 
get_spearman_cor_fixed <- function(data, target, predictors) {
  results <- data.frame(Variable = character(), 
                        Correlation = numeric(), 
                        p_value = numeric(), 
                        stringsAsFactors = FALSE)
  
  for (var in predictors) {
    test <- cor.test(as.numeric(data[[target]]), as.numeric(data[[var]]), 
                     method = "spearman", 
                     use = "pairwise.complete.obs",
                     exact = FALSE)
    
    results <- rbind(results, data.frame(Variable = var, 
                                          Correlation = as.numeric(test$estimate), 
                                          p_value = as.numeric(test$p.value)))
  }
  return(results)
}
```

```{r}
#| include: false
exclude_list <- c("Stage", "Status", "Death", "N_Days", "ID", 
                  "Sex", "Sex_num", "Drug", "Ascites", "Ascites_num", 
                  "Hepatomegaly", "Hepato_num", "Spiders", "Spiders_num", 
                  "Edema", "Edema_num")

continuous_vars <- setdiff(colnames(liver_data), exclude_list)

print(continuous_vars)
```

```{r}
#| echo: false
spearman_res <- get_spearman_cor_fixed(liver_data, "Stage", continuous_vars)
spearman_res$AbsCorr <- abs(spearman_res$Correlation)
top_5_spearman <- spearman_res[spearman_res$p_value < 0.05, ]
top_5_spearman <- 
  top_5_spearman[order(top_5_spearman$AbsCorr, decreasing = TRUE), ][1:5, ]

selected_features <- top_5_spearman$Variable

print("--- [1. Spearman Top 5 Features] ---")
print(top_5_spearman[, c("Variable", "Correlation", "p_value")])

# Conduct multicollinearity diagnostic (VIF)
vif_formula <- as.formula(paste("as.numeric(Stage) ~", 
                                paste(selected_features, collapse = "+")))
vif_model <- lm(vif_formula, data = liver_data)

vif_values <- vif(vif_model)

print("--- [2. Multicollinearity Check (VIF)] ---")
print(vif_values)

max_vif <- max(vif_values)
cat("\n Results: ")
if(max_vif < 5) {
  cat("All VIF values are below 5. Multicollinearity is minimal, and all 5 variables are suitable for inclusion in the model.\n")
} else if(max_vif < 10) {
  cat("Some VIF values exceed 5. While caution is advised, they remain within an acceptable range for analysis.\n")
} else {
  cat("VIF values of 10 or higher detected! Significant multicollinearity is present, and variable refinement may be necessary.\n")
}
```

```{r}
jags_data <- list(
  y = as.numeric(liver_data$Stage),
  alb = as.numeric(scale(liver_data$Albumin)),
  pro = as.numeric(scale(liver_data$Prothrombin)),
  cop = as.numeric(scale(liver_data$Copper)),
  pla = as.numeric(scale(liver_data$Platelets)),
  N = nrow(liver_data)
)

# Define JAGS for an Ordered Probit Model
model_def <- "
model {
  for (i in 1:N) {
    
    # Bayesian Automated Imputation via prior specification
    alb[i] ~ dnorm(mu_alb, tau_alb)
    pro[i] ~ dnorm(mu_pro, tau_pro)
    cop[i] ~ dnorm(mu_cop, tau_cop)
    pla[i] ~ dnorm(mu_pla, tau_pla)
    
    # mu[i] represents the mean of the latent severity probability distribution 
    # for each patient  
    mu[i] <- b_alb*alb[i] + b_pro*pro[i] + b_cop*cop[i] + b_pla*pla[i]
    
    # [Likelihood Part 1: Cumulative Probabilities]
    # Compute the cumulative probabilities using cutpoints (c)  
    # and individual latent severity (mu)
    # Map each patient's linear predictor to the target via the phi function
    p_cum[i, 1] <- phi(c[1] - mu[i]) # Probability of being in stage 1
    p_cum[i, 2] <- phi(c[2] - mu[i]) # Probability of being in stage 1 or 2
    p_cum[i, 3] <- phi(c[3] - mu[i]) # Probability of being in stage 1 or 2 or 3
    
    # [Likelihood Part 2: Individual Category Probabilities]
    # Subtract cumulative probabilities to compute the discrete category 
    # probabilities (pr) for each of the 4 ordinal stages
    pr[i, 1] <- p_cum[i, 1]               # Probability of being in stage 1
    pr[i, 2] <- p_cum[i, 2] - p_cum[i, 1] # Probability of being in stage 2
    pr[i, 3] <- p_cum[i, 3] - p_cum[i, 2] # Probability of being in stage 3
    pr[i, 4] <- 1 - p_cum[i, 3]           # Probability of being in stage 4
    
    # [Final Likelihood Evaluation] 
    # Evaluate the observed ordinal stage (y[i]) for each patient 
    # using the categorical distribution 
    y[i] ~ dcat(pr[i, 1:4])
  }

  # Define priors for coefficients (Uninformative Prior)
  b_alb ~ dnorm(0, 0.001)
  b_pro ~ dnorm(0, 0.001)
  b_cop ~ dnorm(0, 0.001)
  b_pla ~ dnorm(0, 0.001)

  # Define priors for cutpoints (Uninformative Priors with Ordinal Constraints)
  c_raw[1] ~ dnorm(0, 0.001)
  c_raw[2] ~ dnorm(0, 0.001)
  c_raw[3] ~ dnorm(0, 0.001)
  c[1:3] <- sort(c_raw)
  
  # Define hyperparameters for missing data imputation (Uninformative Priors)
  mu_alb ~ dnorm(0, 0.001); tau_alb ~ dgamma(0.001, 0.001)
  mu_pro ~ dnorm(0, 0.001); tau_pro ~ dgamma(0.001, 0.001)
  mu_cop ~ dnorm(0, 0.001); tau_cop ~ dgamma(0.001, 0.001)
  mu_pla ~ dnorm(0, 0.001); tau_pla ~ dgamma(0.001, 0.001)
}
"
```

```{r}
# Configure MCMC setup and execute Bayesian sampling
set.seed(48764388)

RUN_JAGS <- TRUE

if (RUN_JAGS) {

  init_values <- list(
      list(b_alb = 0, b_pro = 0, b_cop = 0, b_pla = 0, 
          c_raw = c(-2.0, 0.0, 2.0), 
          .RNG.name = "base::Wichmann-Hill", .RNG.seed = 123), # Chain 1
      list(b_alb = -0.1, b_pro = 0.1, b_cop = 0.1, 
          b_pla = -0.1, c_raw = c(-1.5, 0.5, 2.5), 
          .RNG.name = "base::Wichmann-Hill", .RNG.seed = 124), # Chain 2
      list(b_alb = 0.1, b_pro = -0.1, b_cop = -0.1, 
          b_pla = 0.1, c_raw = c(-2.5, -0.5, 1.5),
          .RNG.name = "base::Wichmann-Hill", .RNG.seed = 125) # Chain 3
  )

  jags_model <- jags.model(textConnection(model_def), 
                          data = jags_data, 
                          inits = init_values,  
                          n.chains = 3,          
                          n.adapt = 1000)        

  update(jags_model, 2000)

  # Check imputation outcomes together
  samples <- coda.samples(jags_model, 
                          variable.names = c("b_alb", "b_pro", 
                                            "b_cop", "b_pla", "c",
                                            "cop", "pla", "pro"), 
                          n.iter = 5000)

  saveRDS(samples, file = "jags_samples.rds")

  } else {
    if (file.exists("jags_samples.rds")) {
      samples <- readRDS("jags_samples.rds")
      message("Successfully read RDS file.")
    } else {
      stop("Error.")
    }
  }

summary(samples)
```

```{r}
# Check imputation outcome
# [1] 413 (b_cop 1 + patients 412)
all_names <- colnames(samples[[1]])
length(all_names[grep("cop", all_names)])  
length(all_names[grep("pla", all_names)])  
length(all_names[grep("pro", all_names)])
```

```{r}
any(is.na(as.matrix(samples)))
```

```{r}
# Visuaize variable importance
summary_quantiles <- as.data.frame(summary(samples)$quantiles)
summary_stats     <- as.data.frame(summary(samples)$statistics)

target_vars <- c("b_alb", "b_cop", "b_pla", "b_pro")
display_names <- c("Albumin", "Copper", "Platelets", "Prothrombin Time")

plot_data <- data.frame(
  Variable = display_names,
  Mean     = summary_stats[target_vars, "Mean"],
  Lower    = summary_quantiles[target_vars, "2.5%"],
  Upper    = summary_quantiles[target_vars, "97.5%"]
)

ggplot(plot_data, aes(x = Mean, y = reorder(Variable, abs(Mean)))) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 0.8) +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.2, 
                 color = "orange", size = 1) +
  geom_point(size = 3, color = "black") +
  labs(
    title = "Variable Importance (95% Credible Intervals)",
    x = "Estimated Coefficient Value",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12), 
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 10)
  )
```

```{r}
# Visuaize latent thresholds
target_cuts <- c("c[1]", "c[2]", "c[3]")
display_cuts <- c("Cutpoint 1 (Stage 1 | 2)", "Cutpoint 2 (Stage 2 | 3)", "Cutpoint 3 (Stage 3 | 4)")

plot_cuts <- data.frame(
  Variable = display_cuts,
  Mean     = summary_stats[target_cuts, "Mean"],
  Lower    = summary_quantiles[target_cuts, "2.5%"],
  Upper    = summary_quantiles[target_cuts, "97.5%"]
)

ggplot(plot_cuts, aes(x = Mean, y = Variable)) +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.2, 
                 color = "orange", size = 1) +  
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 0.8) +
  geom_point(size = 3, color = "black") +
  labs(
    title = "Latent Thresholds (95% Credible Intervals)",
    x = "Latent Severity Scale",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12), 
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 10)
  )
```

```{r}
#| echo: false
# Draw trace plots
plot(samples[, c("b_alb", "b_cop")])
plot(samples[, c("b_pla", "b_pro")])
plot(samples[, c("c[1]", "c[2]", "c[3]")])
```

```{r}
# Check R_hat and ESS
r_hat_raw <- gelman.diag(samples, multivariate = FALSE)$psrf
r_hat_df  <- as.data.frame(r_hat_raw)

ess_raw <- effectiveSize(samples)
ess_df  <- as.data.frame(ess_raw)

convergence_table <- data.frame(
  Variable = rownames(r_hat_df),
  R_hat    = round(r_hat_df$`Point est.`, 3),    
  ESS      = round(ess_df$ess_raw, 0)             
)

print(convergence_table)
```

According to the Bayesian ordered probit analysis, Albumin, Platelets, Copper, and Prothrombin Time were identified as significant predictors for the histological stages of cirrhosis. Albumin and Platelets exhibit a negative correlation with cirrhosis progression, indicating that a decrease in these biomarkers reflects deteriorating liver function. In contrast, increased levels of Copper and Prothrombin time are associated with advanced cirrhosis stages. While Bilirubin was significant in univariate analysis, it did not show a statistically significant impact in this multivariate model, as its 95% credible interval included zero.

## Research Question 2

What is the most influential blood biomarker for each distinct stage transition?

```{r}
# Identify 

mcmc_summary <- summary(samples)
means_matrix <- mcmc_summary$statistics 
quantiles_matrix <- mcmc_summary$quantiles  

betas <- c(
  Alb = means_matrix["b_alb", "Mean"],
  Cop = means_matrix["b_cop", "Mean"],
  Pla = means_matrix["b_pla", "Mean"],
  Pro = means_matrix["b_pro", "Mean"]
)

raw_data <- data.frame(
  alb   = as.vector(jags_data$alb),
  cop   = as.vector(jags_data$cop),
  pla   = as.vector(jags_data$pla),
  pro   = as.vector(jags_data$pro),
  stage = as.vector(jags_data$y) 
)

if (nrow(raw_data) == 0) {
  stop("Error")
}

mean_stage1 <- colMeans(raw_data[raw_data$stage == 1, 
                                 c("alb", "cop", "pla", "pro")], na.rm = TRUE)
mean_stage2 <- colMeans(raw_data[raw_data$stage == 2, 
                                 c("alb", "cop", "pla", "pro")], na.rm = TRUE)
mean_stage3 <- colMeans(raw_data[raw_data$stage == 3, 
                                 c("alb", "cop", "pla", "pro")], na.rm = TRUE)
names(mean_stage1) <- c("Alb", "Cop", "Pla", "Pro")
names(mean_stage2) <- c("Alb", "Cop", "Pla", "Pro")
names(mean_stage3) <- c("Alb", "Cop", "Pla", "Pro")

calc_stage_share <- function(beta_vec, mean_vec) {
  real_effect <- abs(beta_vec * mean_vec) 
  standardized_shares <- (real_effect / sum(real_effect)) * 100
  return(standardized_shares)
}

share_trans1 <- calc_stage_share(betas, mean_stage1) # Stage 1 -> 2
share_trans2 <- calc_stage_share(betas, mean_stage2) # Stage 2 -> 3
share_trans3 <- calc_stage_share(betas, mean_stage3) # Stage 3 -> 4

get_rankings <- function(share_vector) {
  sorted_shares <- sort(share_vector, decreasing = TRUE)
  return(list(
    Top = paste(names(sorted_shares)[1], " (", round(sorted_shares[1], 1), "%)", sep=""),
    Runner = paste(names(sorted_shares)[2], " (", round(sorted_shares[2], 1), "%)", sep="")
  ))
}

r1 <- get_rankings(share_trans1)
r2 <- get_rankings(share_trans2)
r3 <- get_rankings(share_trans3)

df_real_dominance <- data.frame(
  Transition = c("Stage 1 -> 2 (Early)", "Stage 2 -> 3 (Progression)", 
                 "Stage 3 -> 4 (Severe)"),
  Most_Influential_Biomarker = c(r1$Top, r2$Top, r3$Top),
  Runner_Up = c(r1$Runner, r2$Runner, r3$Runner)
)

print("====================================================================")
print("Mathematical Verification: Dynamic Driver for Each Stage Transition")
print("====================================================================")
print(df_real_dominance)
```

```{r}
df_plot_shares <- data.frame(
  Transition = rep(c("Stage 1 → 2\n(Early)", 
                     "Stage 2 → 3\n(Progression)", 
                     "Stage 3 → 4\n(Severe)"), each = 4),
  Biomarker = rep(c("Albumin", "Copper", 
                    "Platelets", "Prothrombin Time"), times = 3),
  Share = c(
    share_trans1["Alb"], 
    share_trans1["Cop"], 
    share_trans1["Pla"], 
    share_trans1["Pro"],
    share_trans2["Alb"], 
    share_trans2["Cop"], 
    share_trans2["Pla"], 
    share_trans2["Pro"],
    share_trans3["Alb"], 
    share_trans3["Cop"], 
    share_trans3["Pla"], 
    share_trans3["Pro"]
  )
)

df_plot_shares$Transition <- factor(df_plot_shares$Transition, levels = c(
  "Stage 1 → 2\n(Early)", "Stage 2 → 3\n(Progression)", "Stage 3 → 4\n(Severe)"
))

ggplot(df_plot_shares, aes(x = Transition, y = Share, fill = Biomarker)) +
  geom_bar(stat = "identity", 
           position = position_dodge(width = 0.85), 
           width = 0.75) +
  geom_text(aes(label = paste0(round(Share, 1), "%")), 
            position = position_dodge(width = 0.85), 
            vjust = -0.5, size = 2.5, fontface = "bold") +
  
  scale_fill_manual(values = c(
    "Albumin" = "#005088",  
    "Copper" = "#e67e22",   
    "Prothrombin Time" = "#d9534f",  
    "Platelets" = "#94a3b8"   
  )) +
  labs(
    title = "Blood Biomarker Drivers Across Disease Transitions",
    x = "Cirrhosis Stage Transitions",
    y = "Relative Contribution Share (%)",
    fill = "Biomarkers"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    axis.text = element_text(size = 9, face = "bold"),
    axis.title = element_text(size = 10),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank() 
  ) +
  ylim(0, max(df_plot_shares$Share) + 8)
```

Albumin is the most influential driver during the early (stage 1 to 2) and severe (stage 3 to 4) transitions, accounting for 40.3% and 46.1% of the relative contribution share, respectively. In contrast, during the intermediate progression stage (stage 2 to 3), Copper emerges as the dominant driver at 34.1%. Given that Albumin’s influence becomes prominent in the critical severe stage, strict clinical management must focus on maintaining adequate Albumin levels to prevent terminal deterioration.

## Research Question 3

How do clinical changes in blood biomarkers affect stage 4 probability?

```{r}
mcmc_stats <- summary(samples)$statistics

betas <- c(
  Alb = mcmc_stats["b_alb", "Mean"],
  Cop = mcmc_stats["b_cop", "Mean"],
  Pla = mcmc_stats["b_pla", "Mean"],
  Pro = mcmc_stats["b_pro", "Mean"]
)

c_cuts <- c(
  mcmc_stats["c[1]", "Mean"],
  mcmc_stats["c[2]", "Mean"],
  mcmc_stats["c[3]", "Mean"]
)

meta_units <- list(
  Alb = list(
    mean = mean(liver_data$Albumin, na.rm = TRUE),  
    sd   = sd(liver_data$Albumin, na.rm = TRUE),  
    name = "Albumin (g/dL)"
  ),
  Cop = list(
    mean = mean(liver_data$Copper, na.rm = TRUE),  
    sd   = sd(liver_data$Copper, na.rm = TRUE),  
    name = "Copper (µg/dL)"
  ),
  Pla = list(
    mean = mean(liver_data$Platelets, na.rm = TRUE), 
    sd   = sd(liver_data$Platelets, na.rm = TRUE),  
    name = "Platelets (×10³/µL)"
  ),
  Pro = list(
    mean = mean(liver_data$Prothrombin, na.rm = TRUE), 
    sd   = sd(liver_data$Prothrombin, na.rm = TRUE),  
    name = "Prothrombin Time (seconds)"
  )
)

plot_biomarker_real_scale <- function(var_key, coef_value, title_name) {
  
  std_range <- seq(-3, 3, length.out = 100)
  
  probs <- sapply(std_range, function(x) {
    mu <- coef_value * x  
    
    p1 <- pnorm(c_cuts[1] - mu)
    p2 <- pnorm(c_cuts[2] - mu) - pnorm(c_cuts[1] - mu)
    p3 <- pnorm(c_cuts[3] - mu) - pnorm(c_cuts[2] - mu)
    p4 <- 1 - pnorm(c_cuts[3] - mu)
    
    return(c(p1, p2, p3, p4))
  })
  
  var_meta <- meta_units[[var_key]]
  real_range <- (std_range * var_meta$sd) + var_meta$mean
  
  df_plot <- data.frame(
    Value = real_range, 
    `Stage 1` = probs[1,], 
    `Stage 2` = probs[2,], 
    `Stage 3` = probs[3,], 
    `Stage 4` = probs[4,],
    check.names = FALSE 
  )
  
  df_long <- gather(df_plot, key = "Stage", value = "Probability", -Value)
  df_long$Stage <- factor(df_long$Stage, 
                          levels = c("Stage 1", "Stage 2", "Stage 3", "Stage 4"))
  
  p <- ggplot(df_long, aes(x = Value, y = Probability, color = Stage)) +
    geom_line(size = 1.2) +
    labs(title = title_name, x = var_meta$name, y = "Probability") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
      axis.title = element_text(size = 10)
    )
  return(p)
}


p_alb <- plot_biomarker_real_scale("Alb", betas["Alb"], "Albumin (Protective)")
p_cop <- plot_biomarker_real_scale("Cop", betas["Cop"], "Copper (Risk)")
p_pla <- plot_biomarker_real_scale("Pla", betas["Pla"], "Platelets (Protective)")
p_pro <- plot_biomarker_real_scale("Pro", betas["Pro"], "Prothrombin Time (Risk)")

combined_plot <- wrap_plots(p_alb, p_cop, p_pla, p_pro, ncol = 2) + 
  plot_layout(guides = 'collect') + 
  plot_annotation(
    title = "Effect of Blood Biomarkers on Cirrhosis Stage Probabilities (Clinical Units)",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
      legend.title = element_text(face = "bold")
    )
  )

print(combined_plot)
```

```{r}
sd_alb <- sd(liver_data$Albumin, na.rm = TRUE)
sd_cop <- sd(liver_data$Copper, na.rm = TRUE)
sd_pla <- sd(liver_data$Platelets, na.rm = TRUE)
sd_pro <- sd(liver_data$Prothrombin, na.rm = TRUE)

cat("Albumin 1 unit =", round(sd_alb, 2), "g/dL\n")
cat("Copper 1 unit =", round(sd_cop, 2), "ug/dL\n")
cat("Platelets 1 unit =", round(sd_pla, 2), "x 10^3/uL\n")
cat("Prothrombin Time 1 unit =", round(sd_pro, 2), "seconds\n")
```

These four plots illustrate the influence of key blood biomarkers on cirrhosis stage probabilities. Specifically, a one-unit shift in the underlying standardized scale corresponds to a clinical change of $0.42 \text{g/dL}$ for Albumin, $85.61 \mu\text{g/dL}$ for Copper, $95.75 \times 10^3/\mu\text{L}$ for Platelets, and $1.03 \text{ seconds}$ for Prothrombin Time. Notably, stage 4 probability exhibits the most sharp, non-linear progression, suggesting the existence of critical clinical trigger points. The parabolic trend lines of stage 3 intersect with the stage 4 curves at specific thresholds, capturing distinct stage transitions. For example, a Copper level exceeding 150 and a Prothrombin Time rising above 12.5 serve as strong indicators of a transition into stage 4. Conversely, when Albumin drops below 3.3 or Platelets fall below 200, the probability of stage 4 overtakes stage 3, serving as a critical warning sign of transition into the terminal stage.

## Research Question 4

How accurately can the Bayesian Ordered Probit model predict patients in stage 3 and 4?

```{r}
jags_summary <- summary(samples)$statistics

b_alb <- jags_summary["b_alb", "Mean"]
b_cop <- jags_summary["b_cop", "Mean"]
b_pla <- jags_summary["b_pla", "Mean"]
b_pro <- jags_summary["b_pro", "Mean"]

b <- c(b_alb, b_cop, b_pla, b_pro)
c_vals <- jags_summary[c("c[1]", "c[2]", "c[3]"), "Mean"]

scale_alb <- as.numeric(scale(liver_data$Albumin))
scale_cop <- as.numeric(scale(liver_data$Copper))
scale_pla <- as.numeric(scale(liver_data$Platelets))
scale_pro <- as.numeric(scale(liver_data$Prothrombin))

scale_cop[is.na(scale_cop)] <- 0

scale_pla[is.na(scale_pla)] <- 0
scale_pro[is.na(scale_pro)] <- 0

mat_X_perfect <- cbind(scale_alb, scale_cop, scale_pla, scale_pro)

mu_pred <- mat_X_perfect %*% b

N <- nrow(liver_data)
predicted_stage <- integer(N)

for (i in 1:N) {
  p_cum1 <- pnorm(c_vals[1] - mu_pred[i])
  p_cum2 <- pnorm(c_vals[2] - mu_pred[i])
  p_cum3 <- pnorm(c_vals[3] - mu_pred[i])
  
  pr1 <- p_cum1
  pr2 <- p_cum2 - p_cum1
  pr3 <- p_cum3 - p_cum2
  pr4 <- 1 - p_cum3
  
  prob_vector <- c(pr1, pr2, pr3, pr4)
  prob_vector[prob_vector < 0] <- 0  
  
  if (all(prob_vector == 0)) {
    if (mu_pred[i] > 0) { predicted_stage[i] <- 4 } else { predicted_stage[i] <- 1 }
  } else {
    predicted_stage[i] <- which.max(prob_vector)  
  }
}

eval_data <- data.frame(
  Actual = factor(as.numeric(liver_data$Stage), levels = 1:4, 
                  labels = paste("Stage", 1:4)),
  Predicted = factor(predicted_stage, levels = 1:4, labels = paste("Stage", 1:4))
)

conf_matrix <- table(Actual = eval_data$Actual, Predicted = eval_data$Predicted)
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)

print("=== Confusion Matrix ===")
print(conf_matrix)
cat("\n=== Accuarcy ==:\n", round(accuracy * 100, 2), "%\n")

cm_df <- as.data.frame(conf_matrix)

ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b") +
  geom_text(aes(label = Freq), fontface = "bold", size = 5) + 
  labs(title = paste("Final Bayesian Confusion Matrix (Accuracy:", 
                     round(accuracy * 100, 1), "%)"),
       x = "Predicted Stage (Model)",
       y = "Actual Stage (True)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14), 
        axis.title = element_text(face = "bold"))
```

This model serves as a high-safety screening tool. As demonstrated in the final Bayesian confusion matrix, the model achieved 0 cases of critical under-diagnosis, where advanced patients (stage 3 and 4) were never misclassified into stage 1. The overall accuracy of 48.5% reflects the model's preference for cautious over-diagnosis rather than risking fatal under-diagnosis. Notably, the model demonstrates moderate predictive power for advanced conditions, achieving a sensitivity of approximately 68% (106/155) for stage 3 and 56% (81/144) for stage 4. Furthermore, the performance implies that the four standard biomarkers (Albumin, Platelets, Copper and Prothrombin Time) alone are not sufficient to capture individual biological baselines. Lastly, due to the characteristics of the target variables, their boundaries are naturally overlapping, leading to frequent misclassification into adjacent categories (e.g., stage 4 being misclassified as stage 3).
